---
title: "Intro to Spatial Data"
format: html
editor: visual
---
## Download data

```{r}
knb_url <- 'https://dev.nceas.ucsb.edu/knb/d1/mn/v2/object/urn%3Auuid%3Aaceaecb2-1ce0-4d41-a839-d3607d32bb58'
download.file(url = knb_url, destfile = 'demo_data.zip')
unzip('demo_data.zip', exdir = 'data')
file.remove('demo_data.zip')
```
## Set up library
```{r}
library(readr)
library(sf)
library(ggplot2)
library(leaflet)
library(scales)
library(ggmap)
library(dplyr)
```

## Read data


```{r}
ak_regions <- read_sf("data/ak_regions_simp.shp")
```

## Explore data

```{r}
plot(ak_regions)
class(ak_regions)



head(ak_regions)

glimpse(ak_regions)


```
## Coordinate Reference System (CRS)
```{r}
st_crs(ak_regions)
```


## Transform Projection of data

```{r}
ak_regions_3338 <- ak_regions %>%
    st_transform(crs = 3338)

st_crs(ak_regions_3338)

plot(ak_regions_3338)
```


## sf & the Tidyverse

```{r}
colnames(ak_regions_3338)
```
## Using 'select()'

```{r}
ak_regions_3338 %>%
    select(region)
```

## filter() by row

```{r}
unique(ak_regions_3338$region)
```

```{r}
ak_regions_3338 %>%
    filter(region == "Southeast")
```

## Spatial Joins

##  Read in alaska_population.csv using read.csv()

```{r}
# read in population data
pop <- read_csv("data/alaska_population.csv")
```

## Explore data

```{r}
colnames(pop)
head(pop)
glimpse(pop)
class(pop)
```


## Turning DF into sf object

```{r}
pop_4326 <- st_as_sf(pop,
                     coords = c('lng', 'lat'),
                     crs = 4326,
                     remove = F)

head(pop_4326)
class(pop_4326)
```


## Join population data with Alaska regions data using st_join()

```{r}
pop_3338 <- st_transform(pop_4326, crs = 3338)


pop_joined <- st_join(pop_4326, ak_regions_3338, join = st_within)

head(pop_joined)

```

```{r}

pop_3338 <- st_transform(pop_4326, crs = 3338)

pop_joined <- st_join(pop_3338, ak_regions_3338, join = st_within)

head(pop_joined)

```

## Calculate the total population by region using group_by() and summarize()



```{r}
pop_region <- pop_joined %>%
    as.data.frame() %>%
    group_by(region) %>%
    summarise(total_pop = sum(population))

head(pop_region)

```


```{r}
pop_region_3338 <- left_join(ak_regions_3338, pop_region, by = "region")

# plot to check
plot(pop_region_3338)
plot(pop_region_3338["total_pop"])

```
## Calculating population by mgmt area

```{r}
pop_mgmt_338 <- pop_region_3338 %>%
    group_by(mgmt_area) %>%
    summarize(total_pop = sum(total_pop))

plot(pop_mgmt_338["total_pop"])

```
```{r}

pop_mgmt_3338 <- pop_region_3338 %>%
    group_by(mgmt_area) %>%
    summarize(total_pop = sum(total_pop), do_union = F)

plot(pop_mgmt_3338["total_pop"])
```
## Save the spatial object to a new file using write_sf()

```{r}
write_sf(pop_region_3338, "data/ak_regions_population.shp")

```

## Visualization with ggplot

```{r}
ggplot(pop_region_3338) +
    geom_sf(aes(fill = total_pop)) +
    labs(fill = "Total Population") +
    scale_fill_continuous(low = "khaki",
                          high =  "firebrick",
                          labels = comma) +
    theme_bw()

```


